---
title: "Measles"
author: "Amanda Skarlupka"
date: "10/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidytuesdayR)
library(tidyverse)
library(visdat)
library(naniar)
library(readxl)
library(maps)
library(sp)
#library(raster)
#library(mapproj)
library(epitools)
library(stringr)
library(cowplot)
library(ggthemes)
library(Rmisc)
library(dplyr)
```


Measles Vaccination Data
```{r}
tidydata <- tt_load(2020, week=9) 
measles <- tidydata$measles
measles
str(measles)
summary(measles)
```

Measles US Case Data
```{r}
cases <- read_excel("mmc3.xlsx", sheet = "Data_2019 US Measles Cases")
```


First check to see the class and missing data of the variables.
```{r}
vis_dat(measles, warn_large_data = FALSE)
#There is no value of district, and barely any for x-religious exemption. I'm going to remove district, and look into the xreligous.

measles_cl <- measles %>%
  dplyr::select(-district)
```
There seems to be multiple entries for the same school. All of the values are the same but that lat and lng seem to vary a little. First I'll see if the index numbers are unique or if they repeat based on states.

Looking at the vaccination rate data, there are a lot of -1's. I believe that these are NA's and the rate isn't actually negative 1. 

```{r}
measles_cl <- measles_cl %>%
  replace_with_na(replace = list(mmr = -1,
                                 overall = -1))
measles_cl %>%
  filter(index == 1)

# The index is repeated based upon state. So I will group by index, state, name and then take the averages of the lat and longitude.

measles_cl %<>%
  group_by(index, state, name) %>%
  dplyr::summarize(
    type = type,
    city = city,
    county = county,
    enroll = mean(enroll),
    mmr = median(mmr),
    overall = median(overall),
    xmed = median(xmed),
    xper = median(xper),
    lat = mean(lat),
    lng = mean(lng),
            )
measles_cl <- distinct(measles_cl)

vis_dat(measles_cl)
```
Now I want to change the Character strings to factors. 

```{r}
measles_cl$state <- as.factor(measles_cl$state)
measles_cl$county <- as.factor(measles_cl$county)
measles_cl$index <- as.integer(measles_cl$index)
measles_cl$type <- as.factor(measles_cl$type)
levels(measles_cl$type)
vis_dat(measles_cl)
```


I also want to replace nonpublic with private, and BOCES with public. These are publicly funded organizations in New york. 

```{r}


measles_cl$type[measles_cl$type == 'Nonpublic'] <- 'Private'
measles_cl$type[measles_cl$type == 'BOCES'] <- 'Public'


levels(measles_cl$type)
measles_cl$type <- droplevels(measles_cl$type)
levels(measles_cl$type)

```

The enrollment is not well documented. According to the original datafile "Depending on the state, enrollment is for kindergarten only or may extend to include other grades." Therefore I will remove enrollment because I won't be able to tell what that value actually means. 

```{r}

measles_cl <- measles_cl %>%
  dplyr::select(-enroll)

measles_cl$state <- tolower(measles_cl$state)
measles_cl$county <- tolower(measles_cl$county)
```


## Measles Cases Cleaning
```{r}
cases <- cases[,1:4]
cases_orig <- cases

names <- state.name[match(cases$`State Name`,state.abb)]
cases$`State Name` <- names
cases$`State Name` <- as.factor(cases$`State Name`)
vis_dat(cases)
colnames(cases) <- c("subregion", "region", "county_FIPS", "cases")
cases$subregion <- tolower(cases$subregion)
cases$region <- tolower(cases$region)
cases
```
Join the datasets together

```{r}
togeth <- left_join(x = measles_cl, y = cases, by = c("county" = "subregion", "state" = "region"))
vis_dat(togeth)
togeth <- ungroup(togeth)
togeth <- droplevels(togeth)
data <- togeth
```

The data is pretty clean now so lets look at the distributions of the single categorical variables.

Summaries, means, and confidence intervals 
```{r}
summary(togeth$mmr, na.rm = TRUE)
mmr <- togeth %>%
  filter(!is.na(mmr)) %>%
  pull(mmr)
length(mmr)
CI(mmr)

xper <- togeth %>%
  filter(!is.na(xper)) %>%
  pull(xper)
length(xper)
CI(xper)

xmed <- togeth %>%
  filter(!is.na(xmed)) %>%
  pull(xmed)
length(xmed)
CI(xmed)

overall <- togeth %>%
  filter(!is.na(overall)) %>%
  pull(overall)
length(overall)
CI(overall)

```

Cases by US State

```{r}
us_states <- map_data("state")
usa <- map_data("usa")

us_state_case <- left_join(us_states, cases, by = "region")

by_state <- ggplot(data = us_state_case,
            aes(x = long, y = lat,
                group = group, fill = log(cases,10))) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  ggthemes::theme_map() +
  labs(fill = "Log Cases") +
  labs(
    title = "A."
  ) +
  theme(legend.position = "bottom")
fig1_a <- by_state
  
```


Cases by US county

```{r}
county_data <- map_data("county")
county_full <- left_join(county_data, cases, by = c("subregion", "region"))
wash_ore_cases <- cases %>%
  filter(region == "washington" | region == "oregon")
wash_ore_cases_map <- right_join(county_data, wash_ore_cases, by = c("subregion", "region"))

p <- ggplot(data = county_full,
            mapping = aes(x = long, y = lat, fill = log(cases, 10), 
                          group = group))
by_county <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Log Cases") +
  labs(
    title = "B."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig1_b <- by_county

```

Cases by Northwestern US county

```{r}

wash_ore <- county_full %>%
  filter(region == "oregon" | region == "washington")

wash_org_cases <- wash_ore %>%
  filter()

p <- ggplot(data = wash_ore,
            mapping = aes(x = long, y = lat, fill = log(cases, 10), 
                          group = group))
washington_oregon <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Log Cases") +
  labs(
    title = "E."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig2_e <- washington_oregon
fig2_e

```

Cases by Northeastern US county. Not looked into further do to lack of data. 

```{r}
east <- c("new york", "new jersey", "connecticut", "maryland", "massachusetts", "pennsylvania", "rhode island")
eastern <- county_full %>%
  filter(region %in% east)

p <- ggplot(data = eastern,
            mapping = aes(x = long, y = lat, fill = log(cases, 10), 
                          group = group))
eastern_out <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Log Cases") +
  labs(
    title = "D."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
 eastern_out
```


Vaccination data of US and NW by county

```{r}
#Group the vaccination data into county level data. 

county_vaccination <- togeth %>%
  dplyr::select(state, county, mmr, overall, xmed, xper, cases) %>%
  group_by(state, county) %>%
  dplyr::summarize(mmr = median(mmr, na.rm = TRUE),
            overall = median(overall, na.rm = TRUE),
            xmed = median(xmed, na.rm = TRUE),
            xper = median(xper, na.rm = TRUE),
            cases = mean(cases, na.rm = TRUE))


county_vacc <- left_join(county_data, county_vaccination, by = c("subregion" = "county", "region" = "state"))

p <- ggplot(data = county_vacc,
            mapping = aes(x = long, y = lat, fill = mmr, 
                          group = group))
county_mmr <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "MMR vaccination %") +
  labs(
    title = "C."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig1_c <- county_mmr

wash_ore <- county_vacc %>%
  filter(region == "oregon" | region == "washington")

p <- ggplot(data = wash_ore,
            mapping = aes(x = long, y = lat, fill = mmr, 
                          group = group))
wash_ore_mmr <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "MMR vaccination %") +
  labs(
    title = "A."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom") +
  geom_polygon(data = wash_ore_cases_map, color = "red", fill = "grey", alpha = 0)
fig2_a <- wash_ore_mmr

```

```{r}
p <- ggplot(data = county_vacc,
            mapping = aes(x = long, y = lat, fill = xper, 
                          group = group))
county_xper <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Personal Exemption %") +
  labs(
    title = "D."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig1_d <- county_xper

p <- ggplot(data = wash_ore,
            mapping = aes(x = long, y = lat, fill = xper, 
                          group = group))
wash_ore_xper <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Personal Exemption %") +
  labs(
    title = "B."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom") +
  geom_polygon(data = wash_ore_cases_map, color = "red", fill = "grey", alpha = 0)
fig2_b <- wash_ore_xper

```

```{r}
p <- ggplot(data = county_vacc,
            mapping = aes(x = long, y = lat, fill = xmed, 
                          group = group))
county_xmed <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Medical Exemption %") +
  labs(
    title = "E."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig1_e <- county_xmed

p <- ggplot(data = wash_ore,
            mapping = aes(x = long, y = lat, fill = xmed, 
                          group = group))
wash_ore_xmed <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Medical Exemption %") +
  labs(
    title = "C."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom") +
  geom_polygon(data = wash_ore_cases_map, color = "red", fill = "grey", alpha = 0)
fig2_c <- wash_ore_xmed
```

```{r}
p <- ggplot(data = county_vacc,
            mapping = aes(x = long, y = lat, fill = overall, 
                          group = group))
county_overall <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Overall Vaccination %") +
  labs(
    title = "F."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom")
fig1_f <- county_overall

p <- ggplot(data = wash_ore,
            mapping = aes(x = long, y = lat, fill = overall, 
                          group = group))
wash_ore_over <- p +
  geom_polygon(color = "gray90", size = 0.05) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Overall vaccination %") +
  labs(
    title = "D."
  ) +
  ggthemes::theme_map() +
  theme(legend.position = "bottom") +
  geom_polygon(data = wash_ore_cases_map, color = "red", fill = "grey", alpha = 0)
fig2_d <- wash_ore_over

```

Final figures
```{r}
figure1 <- cowplot::plot_grid(fig1_a, fig1_b, fig1_c, fig1_d, fig1_e, fig1_f)
figure2 <- cowplot::plot_grid(fig2_a, fig2_b, fig2_c, fig2_d, fig2_e)

ggsave("figure1.png", plot = figure1, width = 12, height = 10)
ggsave("figure2.png", plot = figure2, width = 10, height = 8)
```

Odds ratio of having a case:

Create factors:

```{r}
northeast <- togeth %>%
  filter(state == "washington" | state == "oregon") %>%
  group_by(state, county) %>%
  dplyr::summarise(mmr = mean(mmr, na.rm = TRUE),
            overall = mean(overall, na.rm = TRUE),
            xmed = mean(xmed, na.rm = TRUE),
            xper = mean(xper, na.rm = TRUE),
            cases = mean(cases, na.rm = TRUE))
northeast <- northeast %>%
  mutate(cases = replace_na(cases, 0))

a <- northeast

quantile(a$mmr, na.rm = TRUE)
quantile(a$xmed, na.rm = TRUE)
quantile(a$overall, na.rm = TRUE)
quantile(a$xper, na.rm = TRUE)

b <- a %>%
  mutate(
    cases = ifelse(cases == 0, "no", "yes"),
    
    mmr_fac = ifelse(mmr < 89, "71-89%", ifelse(mmr <= 92, "89-92%", ifelse(mmr <= 95, "92-95%", "95-100%"))),
    
    xmed_fac = ifelse(xmed < 0.5, "0-0.5%", ifelse(xmed <= 2, "0.5-2.0%", "2-13.7%")),
    
    xper_fac = ifelse(xper < 3, "0.75-3%", ifelse(xper < 5.4, "3-5.4%", ifelse(xper < 7.1, "5.4-7.1%", "7.1-29.2%"))),
    
    overall_fac = ifelse(overall < 83, "53-83%", ifelse(overall <= 89, "83-89%", ifelse(overall <= 93, "89-93%", "93-98%"))),
    
    across(.cols = ends_with("_fac"), as.factor))

northeast_fac <- b

```

Overall Odds Ratio

```{r}
a <- northeast_fac %>%
  filter(!is.na(overall_fac))

b <- table(a$overall_fac, a$cases)

oddsratio(b, rev = "rows")
  
           
```

MMR Odds Ratio

```{r}
a <- northeast_fac %>%
  filter(!is.na(mmr_fac))

b <- table(a$mmr_fac, a$cases)

oddsratio(b, rev = "rows")
  
           
```

Personal Exemption Odds Ratio

```{r}
a <- northeast_fac %>%
  filter(!is.na(xper_fac))

b <- table(a$xper_fac, a$cases)

oddsratio(b)

  
           
```

Medical Exemption Odds Ratio

```{r}
a <- northeast_fac %>%
  filter(!is.na(xmed_fac))

b <- table(a$xmed_fac, a$cases)


oddsratio(b)

riskratio(b)
  
           
```

There is no data on Alaska, Georgia, Hawaii, Indiana, Kentucky, Maryland, Nevada, and New Mexico. 
The data is majority public, then private, kindergarten and finally charter.

The data is pretty clean now so lets look at the distributions of the single continuous variables.

```{r}
togeth %>%
  ggplot(aes(x = overall)) +
  geom_histogram()

togeth %>%
  ggplot(aes(x = overall)) +
  geom_density() +
  facet_wrap(~ type)

togeth %>%
  ggplot(aes(x = overall)) +
  geom_density() +
  facet_wrap(~ state)


togeth %>%
  ggplot(aes(x = mmr)) +
  geom_histogram()

togeth %>%
  ggplot(aes(x = mmr)) +
  geom_density() +
  facet_wrap(~ type)

togeth %>%
  ggplot(aes(x = mmr)) +
  geom_density() +
  facet_wrap(~ state)

togeth %>%
  ggplot(aes(x = xmed)) +
  geom_histogram()

togeth %>%
  ggplot(aes(x = xper)) +
  geom_histogram()

cases %>%
  ggplot(aes(x = cases)) +
  geom_histogram()
  
```








